---
title: "ECNS 560 Project Writeup"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Hide code
  error = FALSE,     # Hide error messages
  warning = FALSE,   # Hide warnings (like "NAs introduced")
  message = FALSE,   # Hide package loading messages
  fig.align = "center" # Centers all your plots automatically
)

library(tidyverse)
library(ggplot2)
library(plotly)
library(yarrr)
library(lubridate)
load("./data/clean/transformed_cleaned_data.RData")
```

## Topic and Motivation

For this project, we aim to predict recessions, with more accuracy than existing indicators. In particular, existing recession indicators tend to be far overfit, and create an indicator as a function of one or two macroeconomic variables which historically change in advance of recessions. These tend to have many false positives -- hence the saying "economists have predicted 11 of the past 5 recessions." We believe we can create an indicator which uses many more variables, and identify a region of space which contains recessions -- when our vector of these variables enters this region, the recession indicator will flag. Importantly, this will rely on all of the variables we use, rather than some small subset.

## Data and Processing

Our data is retrieved from Yahoo Finance (via the "tidyquant" R package) and from FRED (also via the "tidyquant" R package). This data includes time series for the daily close of major stock indices (S&P 500, DOW, NASDAQ, Russel 2000), common hedges for uncertainty (gold, silver, Case-Schiller Index for real estate), and monthly data on macroeconomic data (U3 and U6 unemployment, GDP, CPI, Federal Funds Rate, Housing Starts, ...). Collectively, these present a multi-faceted image of the economy at a point in time -- in particular, they represent both classical macroeconomic indicators and aggregating signals for the "shadow economy" of asset prices and volatility hedges. Further, the vast majority of our data begins well before 2000, with some stretching back to the beginning of the 20th century.

Our primary data of interest is the macroeconomic data -- we are interested in predicting and better understanding recessions, so GDP is one of our most relevant points. GDP is released quarterly, so this would require collapsing much of our data and we would have limited power. Instead, we will primarily use monthly industrial output, which is released monthly, at the same frequency as most of our other macroeconomic data. To match this, we collapse our more granular data (like asset prices) to monthly, creating a monthly mean close price and an interday volatility measure from the variance of these daily closing prices within the month. This collapsed data is combined with the other data, merging on the primary key of year-month, at which point a single primary key for our cleaned dataset is created.

It is also worth noting that there is a "data appendix" in the "clean_data.R" script.

From this, we do some further processing, to create a relative-scaled first-difference dataset. Importantly, this relative-scaled first-difference is computed as a direct first-difference for those values which are already in percentage terms (interest rates or unemployment), and similarly, we also use this direct first-differencing for mean-reverting series, to avoid seeing explosive growth around zero or negligible percent changes for large level-valued changes as it reaches peaks or troughs. Both the transformed and the untransformed data will be useful, but the first difference and relative scaling normalizes and whitens the data, reducing issues of autocorrelation and scale differences between variables.

Finally, we append some functions of our variables onto that dataset to allow for a more robust approach when using persistent homology. In particular, we use log-returns for our asset data and other compounding indices, which tend to exhibit exponential growth and would thus distort the spacial "shape" of the data we intend to exploit in such a manner as to dominate any further processes, despite its failure as a dominant indicator of recessions.

In terms of missing data, as is addressed in the visualizations section of this report, the only missingness of concern is the lack of duration that some features have data for. Aside from that, missing values are of little concern at this stage, keeping in mind our temporal restrictions.

To ensure the robustness and validity of the data, an analysis of outliers is also carried out on the data, using the *dlookr* package in R. Using this tool, we discovered that while there are extreme values, they are naturally occurring, as most can be explained by looking at the historical period in which they lay. This analysis also shed light to the variability of some of our variables, with certain ones being composed of \~35% outlier observations. With all that said, we will keep these extreme values, as not only are they "real observations", but also are vital for the shocks that we are looking for in our Persistent homology approach.

## Visualizations

This plot illustrates what periods we do have data, and where it is missing. As we see here, most of the data is present and once the data series starts, it is almost perfectly complete.

### Data Coverage /Missing Values

```{r}
# 1. Manually list the columns you want to include
vars_to_plot <- c("dow_avg_close", "nasdaq_avg_close", "fedfunds", "gold_price", "housing_starts", "tbill_1yr", "u6_unemployment", "case_shiller")

# 2. Reshape and filter
# Replace 'date_col' with the actual name of your date column
df_long <- altered_data %>%
  pivot_longer(
    cols = all_of(vars_to_plot), 
    names_to = "variable", 
    values_to = "value"
  ) %>%
  # Remove NAs and blanks so the bars stop/start and the axis fits correctly
  filter(!is.na(value) & value != "")

# 3. Create the plot
ggplot(df_long, aes(x = as.Date(date), y = variable)) +
  # Using geom_tile creates the horizontal bars
  # width = 31 ensures the monthly tiles touch to form a solid line
  geom_tile(width = 31, height = 0.7, fill = "steelblue") +
  # expand = c(0,0) forces the axis to start/end exactly at your data points
  scale_x_date(expand = c(0, 0), date_labels = "%b %Y") +
  labs(x = "Time", y = "Variable", title = "Data Availability") +
  theme_minimal()

```

Additionally, while not addressed by this graph, in "merge_data.R", it is revealed that, withstanding a short window where some macroeconomic variables were not recorded (due to the government shutdown in late 2025), there are no missing values that aren't simply due to a series not going that far back.

### 3-D Tracking of points through space

These plots show the path through time the US economy has followed, where color illustrates the time variable, on several three-axis subsets of our data. In particular, we can see that there are some patterns and regions that this path seems to end up in -- our goal is to identify some repeating area where recessions occur, and use that pattern or spacial region to predict future recessions.

```{r}
create_3d_plot <- function(data, x_var, y_var, z_var, title_text) {
  
  # 1. Prep data: drop specific NAs and ensure Date class
  plot_data <- data %>%
    drop_na(all_of(c(x_var, y_var, z_var, "date"))) %>%
    mutate(date = as.Date(date))
  
  # 2. Create numeric scale for the color gradient
  plot_data$date_num <- as.numeric(plot_data$date)
  
  # 3. Build the Time Scale Key (ticks every 5 years)
  date_range <- range(plot_data$date, na.rm = TRUE)
  year_sequence <- seq(year(date_range[1]), year(date_range[2]), by = 5)
  year_numeric <- as.numeric(as.Date(paste0(year_sequence, "-01-01")))

  # 4. Generate Plot
  plot_ly(plot_data, 
          x = ~get(x_var), 
          y = ~get(y_var), 
          z = ~get(z_var), 
          type = 'scatter3d', 
          mode = 'lines',
          line = list(
            color = ~date_num, 
            colorscale = 'Viridis', 
            width = 5,
            colorbar = list(
              title = "Year",
              tickvals = year_numeric,
              ticktext = year_sequence
            )
          )) %>%
    layout(
      title = title_text,
      scene = list(
        xaxis = list(title = x_var),
        yaxis = list(title = y_var),
        zaxis = list(title = z_var)
      ),
      margin = list(l = 0, r = 0, b = 0, t = 40) # Maximize plot area
    )
}

# --- GENERATE INDIVIDUAL PLOTS ---

p1 <- create_3d_plot(altered_data, "dow_avg_close", "dow_volatility", "fedfunds", 
                     "Federal Funds and Dow")

p2 <- create_3d_plot(altered_data, "fedfunds", "case_shiller_logret_scaled", "cpi_logret_scaled", 
                     "Housing, Interest, and Inflation")

p3 <- create_3d_plot(altered_data, "industrial_production_logret", "u6_unemployment", "tbill_10yr", 
                     "Labor and Yield Curve")

p4 <- create_3d_plot(altered_data, "gold_price_logret_scaled", "tbill_1yr", "tbill_10yr", 
                     "Gold and Bonds")

p1
p2
p3
p4
```

### Time series for most variables

Here, we can see how several variable change through time, as well as visualizing when the series starts. This plot allows us to get a better idea of the trends of some of our variables of interest, in order to better inform our understanding of the topological objects created through persistent homology.

```{r}
# "Stacked" time series of returns/percent-changes through time
# For a few financial assets, Industrial Production or whatever that is,
# probably also gold and the Case-Schiller index

selected_vars <- c("dow_avg_close_logret_scaled", "snp500_avg_close_logret_scaled", "tbill_1yr_diff_scaled", "tbill_10yr_diff_scaled", "housing_starts_diff_scaled", "u6_unemployment_diff_scaled", "gold_price_logret_scaled", "case_shiller_logret_scaled", "industrial_production_logret_scaled", "rgdp_logret_scaled")

# Reshape to long format
df_long <- altered_data %>%
  select(date, all_of(selected_vars)) %>%
  pivot_longer(
    cols = -date, 
    names_to = "variable", 
    values_to = "value"
  ) %>%
  # Ensure the order of the panels matches your list above
  mutate(variable = factor(variable, levels = selected_vars))

# Create the stacked panel plot
ggplot(df_long, aes(x = as.Date(date), y = value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_x_date(expand = c(0, 0)) +
  labs(x = "Time", y = NULL) +
  theme_minimal() + # Using minimal to clean up grid lines
  theme(
    # Makes y-labels (variable names) read horizontally
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 8), 
    strip.background = element_blank(),
    strip.placement = "outside",
    # panel.spacing controls the gap between charts
    panel.spacing = unit(0.8, "lines"),
    # This stretches the facets vertically relative to width
    # Lower number = shorter/wider; Higher number = taller
    aspect.ratio = 0.05
  )
```

### Distributions for Variables of Interest

These plots show the distribution of several variables.

```{r, out.width = "100%"}
# 1. Identify numeric columns and grab the first 24
all_numeric_vars <- colnames(altered_data)[sapply(altered_data, is.numeric)]
first_24_vars <- all_numeric_vars[35:52]

# 2. Filter and reshape to long format
df_pirate_long <- altered_data %>%
  select(all_of(first_24_vars)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  # Keep them in the order they appear in your dataframe
  mutate(variable = factor(variable, levels = first_24_vars))

# 3. Create the 3x8 grid
ggplot(df_pirate_long, aes(x = 1, y = value)) +
  # The "Bean" (Density)
  geom_violin(fill = "steelblue", alpha = 0.3, color = "black", size = 0.3) +
  # The "Points" (Raw data)
  geom_jitter(width = 0.1, alpha = 0.1, size = 0.4, color = "black") +
  # The "Bar" (Mean)
  stat_summary(fun = "mean", geom = "crossbar", width = 0.5, color = "red", size = 0.5) +
  # 3x8 Grid
  facet_wrap(~variable, ncol = 6, scales = "free_y") +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text = element_text(size = 7, face = "bold"),
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    subtitle = "Red bar indicates the mean; density 'beans' show distribution shape",
    y = "Value"
  )
```
